Explore tumor cell classification with CellAssign

### 2024-07-23 analyses/cell-type-ewings/exploratory_analysis/01-marker-gene-tumor-classification.Rmd
fix up some of the ewings files with diffs
Clean up notebooks in a few more places, directly grabbing from main to avoid whitespace diffs

# make plot with gene on y-axis, expression on x-axis and then color density by tumor/normal 
# do all marker genes so differences between tumor/normal or only a subset? 
ggplot(celltype_df, aes(x=transformed_gene_expression, color = sum_classification)) +
  geom_density() +
  facet_wrap(vars(gene_symbol)) 
```

@@

# find markers between tumor and normal 
markers <- scran::findMarkers(sce,
                              groups = sce$sum_classification,
                              pval.type = "all") 
```

@@

ggplot(marker_plot_df, aes(x = sum_classification, fill = gene_expresed)) +
  geom_bar(stat = "count") +
  facet_grid(rows = vars(gene_symbol)) +
  scale_fill_manual(values = c("grey", "blue")) +
  theme_classic()

```

@@

# list of heatmaps looking at SingleR/ CellAssign vs tumor/normal 
heatmap <- jaccard_matrices |>
  purrr::imap(
    \(celltype_mat, celltype_method) {
      ComplexHeatmap::Heatmap(
        t(celltype_mat), # transpose because matrix rows are in common & we want a vertical arrangement
        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
        border = TRUE,
        ## Row parameters
        cluster_rows = TRUE,
        row_title = celltype_method,
        row_title_gp = grid::gpar(fontsize = 12),
        row_title_side = "left",
        row_names_side = "left",
        row_dend_side = "right",
        row_names_gp = grid::gpar(fontsize = 10),
        ## Column parameters
        cluster_columns = FALSE,
        column_title = "",
        column_title_gp = grid::gpar(fontsize = 12),
        column_names_side = "bottom",
        column_names_gp = grid::gpar(fontsize = 10),
        column_names_rot = 90,
        ## Legend parameters
        heatmap_legend_param = list(
          title = "Jaccard index",
          direction = "vertical",
          legend_width = unit(1.5, "in")
        ),
        show_heatmap_legend = celltype_method == "SingleR",
      )
    }) |>
  # concatenate vertically into HeatmapList object
  purrr::reduce(ComplexHeatmap::`%v%`) |>
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )
```

@@

- We do see variation in marker gene expression across cells, suggesting that there are both tumor and normal cells here. - Using a z-score cut-off identifies a group of cells that appear to be tumor. .
- The top marker genes for the "tumor" cells contain PRKCB and MAPT, two of the marker genes in our gene list. 
- It's likely that `ADGRG2`, `HES1`, `NROB1`, and `STEAP1` will not be informative in assigning tumor cells. 
- We may be weary of genes like `CD99` too as expression is more consistent across cells rather than specific to a group of cells. 
- Endothelial cells as identified by SingleR and CellAssign generally have low expression of these marker genes and muscle cells have higher expression of marker genes.

>>>>

# make plot with gene on y-axis, expression on x-axis and then color density by tumor/normal
# do all marker genes so differences between tumor/normal or only a subset?
ggplot(celltype_df, aes(x = transformed_gene_expression, color = sum_classification)) +
  geom_density() +
  facet_wrap(vars(gene_symbol))
```

@@

# find markers between tumor and normal
markers <- scran::findMarkers(sce,
  groups = sce$sum_classification,
  pval.type = "all"
)
```

@@

ggplot(marker_plot_df, aes(x = sum_classification, fill = gene_expresed)) +
  geom_bar(stat = "count") +
  facet_grid(rows = vars(gene_symbol)) +
  scale_fill_manual(values = c("grey", "blue")) +
  theme_classic()
```

@@

# list of heatmaps looking at SingleR/ CellAssign vs tumor/normal
heatmap <- jaccard_matrices |>
  purrr::imap(
    \(celltype_mat, celltype_method) {
      ComplexHeatmap::Heatmap(
        t(celltype_mat), # transpose because matrix rows are in common & we want a vertical arrangement
        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
        border = TRUE,
        ## Row parameters
        cluster_rows = TRUE,
        row_title = celltype_method,
        row_title_gp = grid::gpar(fontsize = 12),
        row_title_side = "left",
        row_names_side = "left",
        row_dend_side = "right",
        row_names_gp = grid::gpar(fontsize = 10),
        ## Column parameters
        cluster_columns = FALSE,
        column_title = "",
        column_title_gp = grid::gpar(fontsize = 12),
        column_names_side = "bottom",
        column_names_gp = grid::gpar(fontsize = 10),
        column_names_rot = 90,
        ## Legend parameters
        heatmap_legend_param = list(
          title = "Jaccard index",
          direction = "vertical",
          legend_width = unit(1.5, "in")
        ),
        show_heatmap_legend = celltype_method == "SingleR",
      )
    }
  ) |>
  # concatenate vertically into HeatmapList object
  purrr::reduce(ComplexHeatmap::`%v%`) |>
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )
```

@@

- We do see variation in marker gene expression across cells, suggesting that there are both tumor and normal cells here. - Using a z-score cut-off identifies a group of cells that appear to be tumor. .
- The top marker genes for the "tumor" cells contain `PRKCB` and `MAPT`, two of the marker genes in our gene list. 
- It's likely that `ADGRG2`, `HES1`, `NROB1`, and `STEAP1` will not be informative in assigning tumor cells. 
- We may be weary of genes like `CD99` too as expression is more consistent across cells rather than specific to a group of cells. 
- Endothelial cells as identified by SingleR and CellAssign generally have low expression of these marker genes and muscle cells have higher expression of marker genes.

### 2024-07-23 analyses/cell-type-ewings/template_notebooks/cnv-workflow/02-cellassign.Rmd
fix up some of the ewings files with diffs

```{r}
# function to get assigned cell type based on max prediction returned by CellAssign 
get_celltype_assignment <- function(predictions){

  # if no predictions, assign every cell to NA
  if(length(colnames(predictions)) == 1){
    celltype_assignments <- data.frame(
      barcode = predictions$barcode,
      celltype = NA
    )
  } else {
    # get individual cell type assignments
    # those with the max prediction 
    celltype_assignments <- predictions |>
      tidyr::pivot_longer(
        !barcode,
        names_to = "celltype",
        values_to = "score"
      ) |>
      dplyr::group_by(barcode) |>
      dplyr::slice_max(score, n = 1) |>
      dplyr::ungroup() 
  }

@@

```{r}
# check if marker gene annotations are present 
if(all(is.na(classifications_df$marker_gene_classification))){
  has_marker_gene <- FALSE
  message("No annotations were available using only marker gene expression. 
          Any plots comparing CellAssign to marker gene annotation will be skipped.")
} else {
  has_marker_gene <- TRUE 
}
```

@@

if(all(is.na(tumor_celltype_assignments$celltype))){
  has_tumor_assignments <- FALSE
  message("No CellAssign assignments found for tumor marker genes.")
} else {
 # create a table of assignments 
  table(tumor_celltype_assignments$celltype) 
  has_tumor_assignments <- TRUE
}

```

```{r}
# define annotations to use in all heatmaps
if(has_marker_gene){
  # get annotations for heatmaps
  annotation_df <- data.frame(
    marker_gene_classification = classifications_df$marker_gene_classification,
    row.names = classifications_df$barcodes
  ) 
} else {
  annotation_df <- NULL
}
```

```{r, eval=has_tumor_assignments}
# heatmap of prediction scores 
tumor_predictions |>
  tibble::column_to_rownames("barcode") |> 
  as.matrix() |> 
  pheatmap::pheatmap(show_rownames = FALSE,
                     annotation_row = annotation_df)
```

@@

```{r, eval=has_tumor_assignments}
# make sure that we only run if both classes are found in both cell types 
if(all(c("other", "tumor") %in% celltype_df$cellassign_annotations) &&
   all(c("Tumor", "Normal") %in% celltype_df$marker_gene_classification)){
  # prep annotations for confusion matrix
  celltype_df <- celltype_df |> 
    dplyr::mutate(cellassign_annotations = ifelse(
      # make sure annotations match marker gene annotations
      cellassign_annotations == "tumor", 
      "Tumor", 
      "Normal"
    )) |> 
    # make tumor the positive class
    dplyr::mutate(
      cellassign_annotations = forcats::fct_relevel(cellassign_annotations, "Tumor"),
      marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
    )

  caret::confusionMatrix(
    table(
      celltype_df$marker_gene_classification, 
      celltype_df$cellassign_annotations)
  ) 
}
```

@@

if(all(is.na(visser_celltype_assignments$celltype))){
  has_visser_assignments <- FALSE
  message("No CellAssign assignments found for tumor marker genes.")
} else {
 # create a table of assignments 
  table(visser_celltype_assignments$celltype) 
  has_visser_assignments <- TRUE
}
```

```{r, eval=has_visser_assignments}
# heatmap of prediction scores 
visser_predictions |>
  tibble::column_to_rownames("barcode") |> 
  as.matrix() |> 
  pheatmap::pheatmap(show_rownames = FALSE,
                     annotation_row = annotation_df)
```

@@

# if present, compare across both references 
if(has_endothelial){
  # prep annotations for confusion matrix
  celltype_df <- celltype_df |> 
    dplyr::mutate(
      caret_visser_annotations = ifelse(
        # make sure annotations match marker gene annotations
        visser_cellassign_annotations == "endothelial cells", "Endothelial cells", "Other"
      ),
      caret_cellassign_annotations = ifelse(
        cellassign_celltype_annotation == "Endothelial cells", "Endothelial cells", "Other"
      ),  
      # make tumor the positive class
      caret_visser_annotations = forcats::fct_relevel(caret_visser_annotations, "Endothelial cells"),
      caret_cellassign_annotations = forcats::fct_relevel(caret_cellassign_annotations, "Endothelial cells")
    )

  if(all("Endothelial cells", "Other") %in% celltype_df$visser_cellassign_annotations &&
     all("Endothelial cells", "Other") %in% celltype_df$cellassign_celltype_annotation){
    caret::confusionMatrix(
      table(
        celltype_df$caret_cellassign_annotations, 
        celltype_df$caret_visser_annotations)
    )  
  }
} else {
  message("No endothelial cells were classified by `CellAssign`.")
}
```

@@

    # faceted umap showing a umap panel for each marker gene 
    ggplot(plot_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
      geom_point(alpha = 0.1, size = 0.2) +
      facet_wrap(vars(gene_symbol)) +
      scale_color_viridis_c() +
      labs(
        title = celltype,
        color = "Log-normalized gene expression"
      ) +
      # remove axis numbers and background grid
      scale_x_continuous(labels = NULL, breaks = NULL) +
      scale_y_continuous(labels = NULL, breaks = NULL) +
      theme(
        aspect.ratio = 1,
        legend.position = "bottom",
        axis.title = element_text(size = 9, color = "black"),
        strip.text = element_text(size = 8),
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 8)
      ) +
      guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) +
      theme_bw()

  })
```

@@

if(all(is.na(panglao_celltype_assignments$celltype))){
  has_panglao_assignments <- FALSE
  message("No CellAssign assignments found for tumor marker genes.")
} else {
 # create a table of assignments 
  table(panglao_celltype_assignments$celltype) 
  has_panglao_assignments <- TRUE
}
```

```{r, eval=has_panglao_assignments}
# heatmap of prediction scores 
panglao_predictions |>
  tibble::column_to_rownames("barcode") |> 
  as.matrix() |> 
  pheatmap::pheatmap(show_rownames = FALSE,
                     annotation_row = annotation_df)
```

@@

# umap showing cell type annotations from using Visser reference 
ggplot(celltype_df, aes(x = UMAP1, UMAP2, color = panglao_cellassign_annotations)) +
  geom_point(size = 0.5, alpha = 0.5) + 
  theme_bw()

```

@@

```{r}
celltype_df <- celltype_df |> 
  dplyr::select(barcodes, 
                marker_gene_cellassign_annotations = cellassign_annotations, 
                visser_cellassign_annotations, 
                panglao_cellassign_annotations)

>>>>

```{r}
# function to get assigned cell type based on max prediction returned by CellAssign
get_celltype_assignment <- function(predictions) {
  # if no predictions, assign every cell to NA
  if (length(colnames(predictions)) == 1) {
    celltype_assignments <- data.frame(
      barcode = predictions$barcode,
      celltype = NA
    )
  } else {
    # get individual cell type assignments
    # those with the max prediction
    celltype_assignments <- predictions |>
      tidyr::pivot_longer(
        !barcode,
        names_to = "celltype",
        values_to = "score"
      ) |>
      dplyr::group_by(barcode) |>
      dplyr::slice_max(score, n = 1) |>
      dplyr::ungroup()
  }

@@

```{r}
# check if marker gene annotations are present
if (all(is.na(classifications_df$marker_gene_classification))) {
  has_marker_gene <- FALSE
  message("No annotations were available using only marker gene expression.
          Any plots comparing CellAssign to marker gene annotation will be skipped.")
} else {
  has_marker_gene <- TRUE
}
```

@@

if (all(is.na(tumor_celltype_assignments$celltype))) {
  has_tumor_assignments <- FALSE
  message("No CellAssign assignments found for tumor marker genes.")
} else {
  # create a table of assignments
  table(tumor_celltype_assignments$celltype)
  has_tumor_assignments <- TRUE
}
```

```{r}
# define annotations to use in all heatmaps
if (has_marker_gene) {
  # get annotations for heatmaps
  annotation_df <- data.frame(
    marker_gene_classification = classifications_df$marker_gene_classification,
    row.names = classifications_df$barcodes
  )
} else {
  annotation_df <- NULL
}
```

```{r, eval=has_tumor_assignments}
# heatmap of prediction scores
tumor_predictions |>
  tibble::column_to_rownames("barcode") |>
  as.matrix() |>
  pheatmap::pheatmap(
    show_rownames = FALSE,
    annotation_row = annotation_df
  )
```

@@

```{r, eval=has_tumor_assignments}
# make sure that we only run if both classes are found in both cell types
if (all(c("other", "tumor") %in% celltype_df$cellassign_annotations) &&
  all(c("Tumor", "Normal") %in% celltype_df$marker_gene_classification)) {
  # prep annotations for confusion matrix
  celltype_df <- celltype_df |>
    dplyr::mutate(cellassign_annotations = ifelse(
      # make sure annotations match marker gene annotations
      cellassign_annotations == "tumor",
      "Tumor",
      "Normal"
    )) |>
    # make tumor the positive class
    dplyr::mutate(
      cellassign_annotations = forcats::fct_relevel(cellassign_annotations, "Tumor"),
      marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
    )

  caret::confusionMatrix(
    table(
      celltype_df$marker_gene_classification,
      celltype_df$cellassign_annotations
    )
  )
}
```

@@

if (all(is.na(visser_celltype_assignments$celltype))) {
  has_visser_assignments <- FALSE
  message("No CellAssign assignments found for tumor marker genes.")
} else {
  # create a table of assignments
  table(visser_celltype_assignments$celltype)
  has_visser_assignments <- TRUE
}
```

```{r, eval=has_visser_assignments}
# heatmap of prediction scores
visser_predictions |>
  tibble::column_to_rownames("barcode") |>
  as.matrix() |>
  pheatmap::pheatmap(
    show_rownames = FALSE,
    annotation_row = annotation_df
  )
```

@@

# if present, compare across both references
if (has_endothelial) {
  # prep annotations for confusion matrix
  celltype_df <- celltype_df |>
    dplyr::mutate(
      caret_visser_annotations = ifelse(
        # make sure annotations match marker gene annotations
        visser_cellassign_annotations == "endothelial cells", "Endothelial cells", "Other"
      ),
      caret_cellassign_annotations = ifelse(
        cellassign_celltype_annotation == "Endothelial cells", "Endothelial cells", "Other"
      ),
      # make tumor the positive class
      caret_visser_annotations = forcats::fct_relevel(caret_visser_annotations, "Endothelial cells"),
      caret_cellassign_annotations = forcats::fct_relevel(caret_cellassign_annotations, "Endothelial cells")
    )

  if (all("Endothelial cells", "Other") %in% celltype_df$visser_cellassign_annotations &&
    all("Endothelial cells", "Other") %in% celltype_df$cellassign_celltype_annotation) {
    caret::confusionMatrix(
      table(
        celltype_df$caret_cellassign_annotations,
        celltype_df$caret_visser_annotations
      )
    )
  }
} else {
  message("No endothelial cells were classified by `CellAssign`.")
}
```

@@

    # faceted umap showing a umap panel for each marker gene
    ggplot(plot_df, aes(x = UMAP1, y = UMAP2, color = gene_expression)) +
      geom_point(alpha = 0.1, size = 0.2) +
      facet_wrap(vars(gene_symbol)) +
      scale_color_viridis_c() +
      labs(
        title = celltype,
        color = "Log-normalized gene expression"
      ) +
      # remove axis numbers and background grid
      scale_x_continuous(labels = NULL, breaks = NULL) +
      scale_y_continuous(labels = NULL, breaks = NULL) +
      theme(
        aspect.ratio = 1,
        legend.position = "bottom",
        axis.title = element_text(size = 9, color = "black"),
        strip.text = element_text(size = 8),
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 8)
      ) +
      guides(colour = guide_colorbar(title.position = "bottom", title.hjust = 0.5)) +
      theme_bw()
  })
```

@@

if (all(is.na(panglao_celltype_assignments$celltype))) {
  has_panglao_assignments <- FALSE
  message("No CellAssign assignments found for tumor marker genes.")
} else {
  # create a table of assignments
  table(panglao_celltype_assignments$celltype)
  has_panglao_assignments <- TRUE
}
```

```{r, eval=has_panglao_assignments}
# heatmap of prediction scores
panglao_predictions |>
  tibble::column_to_rownames("barcode") |>
  as.matrix() |>
  pheatmap::pheatmap(
    show_rownames = FALSE,
    annotation_row = annotation_df
  )
```

@@

# umap showing cell type annotations from using Visser reference
ggplot(celltype_df, aes(x = UMAP1, UMAP2, color = panglao_cellassign_annotations)) +
  geom_point(size = 0.5, alpha = 0.5) +
  theme_bw()
```

@@

```{r}
celltype_df <- celltype_df |>
  dplyr::select(barcodes,
    marker_gene_cellassign_annotations = cellassign_annotations,
    visser_cellassign_annotations,
    panglao_cellassign_annotations
  )
